Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild

### 2020-10-08 IJHCSSubmission/bp_studyMethod.tex
Addressing the reviewers last minor comments

Keeping participants engaged in an eight-week study can be
difficult. In the second month of the study, largely to incent participants to
continue, we offered participants two one-hour Tai Chi classes per
week (each in the middle of the day) and asked them to attend one
class a week. The choice of Tai Chi provided a link to mindfulness,
a topic of growing interest in communities. Consultations we held
with researchers in psychiatry and psychology suggested Tai Chi as
a good link to mindfulness and an intervention that might alleviate
stress. By recording attendance of participants at the Tai Chi
sessions, we are able to analyze the impact of the sessions
on the results. As a brief summary, the impact of the
sessions was minimal; we present an analysis
of the effects in Section~\ref{sec:Discussion}.

@@

\subsection{Data Collection}
\label{sec:DataCol}
We collected two datasets  from each study participant.

@@

Table~\ref{signals} lists the biometrics measurements that we collected using the Everion. Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating. Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study. We chose these biometric measurements based on previous research that has indicated their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.

@@

\subsubsection{Surveys}
\label{sec:Surveys}
Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
following the preferences of extensive user piloting, we sent via 
text message a survey request to each participant two times per 
work day. Pilot participants preferred the usage of text message, in part, due 
to them being accessible and noticeable anywhere in the office.

We sent the 
first request at a random time between 9am and 11am 
and sent the second request at a random time between 1pm and 3pm. We 
randomized the request times to avoid either establishing or observing a 
standard behavioral pattern. That is, we did not want the participants to 
plan for the arrival of the survey request at a set time, and we did not 
want the survey request to overlap with a set daily behavior (e.g., coffee 
break every day at 2:30pm). Similarly, we avoided using tools which allow 
for too much freedom in response time~\cite{Adams18} since this would 
discourage participation in stressed time frames and would bias the 
corpus. 
%This sentence was added in response to Reviewer 1 comment:
%There is already a lot of work on designing a new self-log, a self-report 
%tool which can blend into the workstation [1] or offers users more freedom 
%in time to self-report [2], or facilitate the self-log by using wearables.
The same survey was sent each time:
\begin{enumerate}
\item How awake are you right now?
\item How stressed do you feel right now?
\item How focused on work are you right now? 
\end{enumerate}
We used the phrase ``right now'' to capture each aspect in the moment (so as to permit later prediction of each aspect based on biometric data). The wordings of the questions are based on a previous survey of individuals in an organizational context~\cite{Gloor_etal:2010}. The use of awakeness (rather than sleepiness) in Question 1 is inspired by previous work~\cite{Wilhelm_Schoebi:2007} and to some extent also captures the ``arousal'' aspect of the affective space~\cite{Russell:1980}.

%\todo{we are not using the last survey for anything in this paper, so I'd leave it out}
One last survey was sent at the end of the day, at 4:25pm which asked the 
four different questions
detailed below:
\begin{enumerate}
\item How awake have you been today?
\item How stressed did you feel today?
\item How productive have you been today?
\item How do you feel about your work day?
\end{enumerate}

@@

We did not ask our participants about their focus levels in the end of the day survey as focus
is more of an in-the-moment aspect than stress and awakeness.

>>>>

Keeping participants engaged in an eight-week study can be
difficult. In the second month of the study, largely to \rev{incentivize} participants to
continue, we offered participants two one-hour Tai Chi classes per
week (each in the middle of the day) and asked them to attend one
class a week. The choice of Tai Chi provided a link to mindfulness,
a topic of growing interest in communities. Consultations we held
with researchers in psychiatry and psychology suggested Tai Chi as
a good link to mindfulness and an intervention that might alleviate
stress. By recording attendance of participants at the Tai Chi
sessions, we are able to analyze the impact of the sessions
on the results. As a brief summary, the impact of the
sessions was minimal; we present an analysis
of the effects in Section~\ref{sec:Discussion}.

@@

\subsection{Data Collection}
\label{sec:DataCol}
We collected \rev{three} datasets  from each participant \rev{as described below:}

@@

Table~\ref{signals} lists the biometrics measurements that we collected using the Everion. Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating. Data collected by the Everion \rev{is} uploaded to a server, from which we downloaded the data for use in our study. We chose these biometric measurements based on previous research \rev{which indicates} their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.

@@

\subsubsection{Surveys}
\label{sec:Surveys}
Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
following the preferences of extensive user piloting, we sent via 
text message a survey request to each participant two times per 
workday. Pilot participants preferred the usage of text message, in part, due 
to them being accessible and noticeable anywhere in the office.

We sent the 
first request at a random time between 9am and 11am 
and sent the second request at a random time between 1pm and 3pm. We 
randomized the request times to avoid either establishing or observing a 
standard behavioral pattern. That is, we did not want the participants to 
plan for the arrival of the survey request at a set time, and we did not 
want the survey request to overlap with a set daily behavior (e.g., coffee 
break every day at 2:30pm). Similarly, we avoided using tools which allow 
for too much freedom in response time~\cite{Adams18} since this would 
discourage participation in stressed time frames and would bias the 
corpus. 
%This sentence was added in response to Reviewer 1 comment:
%There is already a lot of work on designing a new self-log, a self-report 
%tool which can blend into the workstation [1] or offers users more freedom 
%in time to self-report [2], or facilitate the self-log by using wearables.
The same survey was sent each time \rev{consisting of the following three questions}:
\begin{enumerate}
\item How awake are you right now?
\item How stressed do you feel right now?
\item How focused on work are you right now? 
\end{enumerate}
We used the phrase ``right now'' to capture each aspect in the moment (so as to permit later prediction of each aspect based on biometric data). The \rev{wording of the questions is} based on a previous survey of individuals in an organizational context~\cite{Gloor_etal:2010}. The use of awakeness (rather than sleepiness) in Question 1 is inspired by previous work~\cite{Wilhelm_Schoebi:2007} and to some extent also captures the ``arousal'' aspect of the affective space~\cite{Russell:1980}.

%\todo{we are not using the last survey for anything in this paper, so I'd leave it out}
One last survey was sent at the end of the day at 4:25pm\rev{,} which asked the 
four different questions
detailed below:
\begin{enumerate}
\item How awake have you been today?
\item How stressed did you feel today?
\item How productive have you been today?
\item How do you feel about your workday?
\end{enumerate}

@@

We did not ask our participants about their focus levels in the end of the day survey\rev{,} as focus
is more of an in-the-moment aspect than stress and awakeness.

### 2020-10-08 IJHCSSubmission/bp_threats.tex
Addressing the reviewers last minor comments

\subsection{External Validity}
The results of our study may 
 not generalize to a broader population of office workers.
To mitigate this risk, we collected participants
from a wide variety of different departments
with different age ranges, genders, work experience, and 
working in different positions.

Another threat is that our results may not generalize
to a different office environment. We conducted
this study in a typical office environment, similar to many
among technology workers across the world.
These office environments control for a series of
variables to make them standard worldwide such
as controlled temperature and lighting.

>>>>

\subsection{External Validity}
The results of our study may 
 not generalize to a broader population of office workers.
To mitigate this risk, we collected participants
from a wide variety of departments
with different age ranges, genders, work experience, and 
working in different positions \rev{therefore providing evidence
that our approach's performance is comprised of and can generalize to a wide range of 
knowledge workers}.

\rev{Secondly,} our results may not generalize
to a different office environment. We conducted
this study in a typical office environment, similar to many
among technology workers across the world.
These office environments control for a series of
variables to make them standard worldwide such
as controlled temperature and lighting.

### 2020-10-08 IJHCSSubmission/elsarticle-template.tex
Addressing the reviewers last minor comments

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
   \documentclass[final,1p,times,plainnat]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

@@

\begin{abstract}
Knowledge workers face many challenges in the workplace: work is 
fragmented, disruptions are constant, tasks are complex, and work hours can be 
long. These challenges can impact knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
performed and their productivity in general. We report on a field study with 14
knowledge workers over an eight-week period in which we investigated, using
experience sampling, how the workers experience stress \rev{and} awakeness over time. During this field study, we also collected biometric data including heart- and skin-related measures, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
We observed and report on various trends in knowledge worker stress and awakeness levels
over several weeks, finding that people tend to have certain ``baseline'' levels for these aspects\rev{. Moreover, we found that days with high levels of stress tend to cluster, similarly as the days with low awakeness}. We further show that machine learning models can be built
from the data of a single minimally invasive \rev{device} to
predict stress, focus, and awakeness. Overall, we found that our models were capable of large improvements in precision and recall in comparison to a random classifier for stress (25.9\% increase over random for precision, 4.2\% for recall) and awakeness (52.4\% increase in precision, 40.8\% in recall). The abstract concept of focus proved to be the hardest to predict (26.0\% increase in precision, 27.8\% decrease in recall).
\end{abstract}

>>>>

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
   \documentclass[final, 1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

@@

\begin{abstract}
Knowledge workers face many challenges in the workplace: work is 
fragmented, disruptions are constant, tasks are complex, and work hours can be 
long. These challenges can \rev{affect} knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
performed and their productivity in general. We report on a field study with 14
knowledge workers over an eight-week period in which we investigated, using
experience sampling, how the workers experience stress \rev{and} awakeness over time. During this field study, we also collected biometric data including heart- and skin-related measures, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
We observed and report on various trends in knowledge worker stress and awakeness levels
over several weeks, finding that people tend to have certain ``baseline'' levels for these aspects\rev{. Moreover, we found that days with high levels of stress tend to cluster, similarly as the days with low awakeness}. We further show that machine learning models can be built
from the data of a single minimally invasive \rev{device} to
predict stress, focus, and awakeness. Overall, we found that our models were capable of large improvements in precision and recall in comparison to a random classifier for stress (25.9\% increase over random for precision, 4.2\% for recall) and awakeness (52.4\% increase in precision, 40.8\% in recall). The abstract concept of focus proved to be the hardest to predict (26.0\% increase in precision, 27.8\% decrease in recall).
\end{abstract}
